What is happening on Twitter? A framework for student research projects with tweets

### 2020-09-06 Rmd/reply-to-reviewers.Rmd

> I have only two general suggestions: first, it would be nice to have a brief description of the two undergraduate projects within the manuscript. While I understand the data is no longer available, and think it was the right decision to focus on the case study for methods and results, including a short paragraph about each of the undergraduate projects you supervised would help to give the reader a sense of the range of possibilities of a twitter-based project. This is briefly mentioned at the end of the manuscript but it would help to see the motivations for such a project earlier.

> Second, while I appreciate the concise and clear writing, adding transitions between the sections would help the reader anticipate what is coming. As one example, prior to Sec 5.2, there is no indication that the authors will walk through (a) querying twitter API (b) accessing full tweets via ID numbers (c) parsing JSON.

> Specific Comments:

> - Page 2 Line 28: Should there be a reference for “rhetoric in recruiting political supporters”?

**Yes, we mistakenly omitted them. We've now added key references for this point.**

> - Page 2 Line 52: I’d recommend adding a statement along the lines of: “different statistics and data science researchers may categorize these skills differently, but our assignment is based on our program/specific projects.” For example, I imagine some projects using twitter data may not categorize “Use text analysis tools to analyze tweets” as “Enduring Understanding”, but would categorize “Use data visualization to clarify and inform quantitative analysis” in such a way.

**We've now added a statement like this to the manuscript to reflect this point.**

> - Page 4 Sec 4.2: I agree with the authors that differences in student research interests and goals should be an important consideration and it would be nice to see some examples.

**We've added two examples from our students.**

> - Page 4 Line 82: Would this type of project also be appropriate for a summer research experience?
> - Page 4 Line 88-92: This wording could be a little more clear: is the case study essentially a replicate of a student project? It would be helpful to see an overview of what the original project was, and then introduce the case study as a replicate.
> - Page 5 Sec 5.2: would this process theoretically work for both windows and Mac machines? Is it necessary to use a crontab-like process, or could it be done entirely within R?

@@

> - Page 7 Line 133: Include citations for rtweet and tidy text appear here, where they’re mentioned first, instead of in the following section.

**We've now moved the citations to the first mention of these packages.**

> - Page 8 Line 152: It would be helpful to see a brief model equation.

**We agree! We've added the equations that describe Latent Dirichlet Allocation.**

### 2020-09-06 Rmd/jse-special-issue-2020-october.Rmd

Studies of Twitter conversations have yielded valuable insights into 
modern culture. Using large collections of tweets, scholars have 
investigated diverse research questions, including
the inference of relationships and social networks among Twitter users [@lin2011joint]; 
authorship of specific tweets when multiple persons share a single account [@drob]; and 
rhetoric in recruiting political supporters [@pelled2018little;@wells2016trump].
Recognizing the potential utility of tweets for data science research and teaching, 
we created a collection of tweets over time by repeated querying of the Twitter streaming API.

@@

# Backward design and learning objectives

@@

Backward design principles guided our planning and informed the writing of
learning objectives [@wiggins2005understanding]. Following @wiggins2005understanding, 
we began by listing 
what students, at the end of their thesis research, should be able to do, 
understand, and know. We then classified each of these items into 
one of three categories: enduring understanding, important to know and do, 
and worth being familiar with [@wiggins2005understanding] (Table \@ref(tab:circle-table)). 
Nearly all of the skills in Table \@ref(tab:circle-table) are 
transferable. They apply not merely to thesis projects, but also to
data science research in general.

```{r, circle-table, echo = FALSE, message=FALSE}
library(magrittr)
readr::read_csv("../data/concepts-backward-design.csv") %>%
  knitr::kable(caption = "Classifying project skills")
```

@@

We translatedour prioritized list of skills that 
students should be able to do, understand, and know into learning objectives (Table \@ref(tab:circle-table)). 
We phrased learning objectives in a manner that enabled their subsequent assessment 
(Table \@ref(tab:summative)) with formative and summative strategies. These were 
our four learning objectives:

@@

# Preliminary research mentoring considerations

@@

One of our two students had extensive experience in statistical computing. In 
addition to R computing skills, she also worked in Python and excelled in shell scripting. 
She first learned Python in computer science courses. Our second student had 
extensive experience with R from his statistics courses. His background 
enabled him to write an R package as part of his senior project.

Our data science projects required that students possess elementary R computing
skills. We guided students towards supplementary R computing resources, including 
the online books "R for Data Science" [@wickham2016r] and "Advanced R" [@wickham2019advanced].

@@

We briefly describe the two student projects to give readers a better sense of research possibilities with tweets. Our first student examined relationships over time between 
stock market index prices and tweet sentiment. For each day in her 12-month study period, 
she identified stock market-related tweets with a key word search. With the complete
texts of stock market-related tweets for each day, she calculated a daily sentiment score and plotted it over time. 
Her sentiment score reflected presence of emotion-associated terms (eg., "happy", "sad", "mad", "scared") 
in tweet texts. Days with more net positive emotion words in the collected tweets received a higher (positive) daily sentiment score, while days with more net negative words received a negative daily sentiment score.

For her final project, she presented plots over time of her daily sentiment scores and daily closing prices of the Standard and Poor's 500 index. She also explored time series analysis methods to quantify relationships between closing index prices and sentiment scores.

@@

Our two statistics students conducted their research projects during the 2015-2016 academic year. 
We recommend a full academic year for projects of this magnitude, although a one-semester 
project is possible. Our students presented their findings at the statistics department’s 
undergraduate poster session near the end of the 2015-2016 academic year (Supplementary files). We present below reproducible R code for analyzing data from May 2020. While these
are not the same data that our students analyzed in 2015, the methods and code 
are very similar.

@@

>>>>

Studies of Twitter conversations have yielded valuable insights into 
modern culture. Using large collections of tweets, scholars have 
investigated diverse research questions, including
the inference of relationships and social networks among Twitter users [@lin2011joint]; 
authorship of specific tweets when multiple persons share a single account [@drob]; and 
rhetoric in recruiting political supporters [@pelled2018little;@wells2016trump].
Recognizing the potential utility of tweets for data science research and teaching, 
we created a collection of tweets over time by repeated querying of the Twitter streaming API. 
We envisioned this collection as a rich resource for data science research projects. This vision
grew into two mentored undergraduate student research projects in the 2015-2016 academic year.

@@

# Structure of mentored research

@@

Backward design principles guided our planning and informed the writing of
learning objectives [@wiggins2005understanding]. Following @wiggins2005understanding, 
we began by listing 
what students, at the end of their thesis research, should be able to do, 
understand, and know. We then classified each of these items into 
one of three categories: enduring understanding, important to know and do, 
and worth being familiar with [@wiggins2005understanding] (Table \@ref(tab:circle-table)). 
While other researchers may categorize these skills differently, 
our assignments reflect our projects' priorities.
Nearly all of the skills in Table \@ref(tab:circle-table) are 
transferable. They apply not merely to thesis projects, but also to
data science research in general.

```{r, circle-table, echo = FALSE, message=FALSE}
library(magrittr)
readr::read_csv("../data/concepts-backward-design.csv") %>%
  knitr::kable(caption = "Classifying project skills") %>%
  kableExtra::column_spec(column = 1, border_left = TRUE) %>%
  kableExtra::column_spec(column = 2, border_right = TRUE)
```

@@

We translated our prioritized list of skills that 
students should be able to do, understand, and know into learning objectives (Table \@ref(tab:circle-table)). 
We phrased learning objectives in a manner that enabled their subsequent assessment 
(Table \@ref(tab:summative)) with formative and summative strategies. These were 
our four learning objectives:

@@

## Preliminary research mentoring considerations

@@

One of our two students had extensive experience in statistical computing. In 
addition to R computing skills, she also worked in Python and excelled in shell scripting. 
She first learned Python in computer science courses. Our second student had 
extensive experience with R from his statistics courses. His background 
enabled him to write an R package as part of his senior project. To encourage 
further development of R computing skills in our two students, we guided them 
towards the online books "R for Data Science" [@wickham2016r] and "Advanced R" [@wickham2019advanced].

@@

We briefly describe the two student projects to give readers a better sense 
of research possibilities with tweets. Our first student examined relationships over time between 
stock market index prices and tweet sentiment. For each day in her 12-month study period, 
she identified stock market-related tweets with a key word search. With the complete
texts of stock market-related tweets for each day, she calculated a daily sentiment score and plotted it over time. 
Her sentiment score reflected presence of emotion-associated terms 
(*eg*., "happy", "sad", "mad", "scared") 
in tweet texts. Days with more net positive emotion words in the collected tweets received a higher 
(positive) daily sentiment score, while days with more net negative words received a negative daily 
sentiment score. 
For her final project, she presented plots over time of her daily sentiment scores and daily closing 
prices of the Standard and Poor's 500 index. She also explored time series analysis methods to quantify 
relationships between index prices and sentiment scores.

Our second student developed social media event detection methods with topic 
models. He hypothesized that tweet content changes over time, and that we 
might detect these changes by comparing inferred tweet topics from distinct 
time periods. To validate his hypothesis, he examined tweet content before, during, and after 
the National Football League's Super Bowl game in 2015. He reasoned that 
because the Super Bowl is widely discussed on Twitter, we might detect 
Super Bowl-related topics from tweets sent during the game, but that the football-related
topics would be short-lived in the continuous Twitter stream. We discovered
evidence to support his strategy, and we ultimately presented our findings 
at international and local research meetings. We share below a case study based
on this approach.

@@

Our two statistics students conducted their research projects during the 2015-2016 academic year. 
We recommend a full academic year for projects of this magnitude, although a summer or one-semester 
project is possible. Our students presented their findings at the statistics department’s 
undergraduate poster session near the end of the 2015-2016 academic year (Supplementary files). We 
present below reproducible R code for analyzing data from May 2020. While these
are not the same data that our students analyzed in 2015, the methods and code 
are very similar to that of our second student's project.

@@

We also note that the murder of George Floyd occurred on May 25. Our last 
examined time period, from 12:00 pm to 12:05 pm (Eastern USA time zone) on May 26, occurred after Floyd's murder, 
yet we didn't detect this event in our ten-topic LDA model. Several considerations 
may account for this. While outrage at the murder eventually spread worldwide, there may have been few 
Floyd-related tweets during our collection time on May 26, less than 24 hours after the murder and video release. 
Had we extended our analysis to May 27 and beyond, we may have identified George Floyd-related topics.

### 2020-09-06 Rmd/student-projects.Rmd

One student examined relationships between 
stock market index prices and tweet sentiment. For each day in her 12-month study period, 
she identified stock market related tweets with a key word search. With the complete
texts of stock market-related tweets for each day, she calculated a daily sentiment score and plotted it over time. 
Her sentiment score reflected presence of emotion-associated terms (eg., "happy", "sad", "mad", "scared") 
in tweet texts. Days with more net positive emotion words in the collected tweets received a higher (positive) daily sentiment score, while days with more net negative words received a negative daily sentiment score.

>>>>

One student examined relationships over time between 
stock market index prices and tweet sentiment. For each day in her 12-month study period, 
she identified stock market-related tweets with a key word search. With the complete
texts of stock market-related tweets for each day, she calculated a daily sentiment score and plotted it over time. 
Her sentiment score reflected presence of emotion-associated terms (eg., "happy", "sad", "mad", "scared") 
in tweet texts. Days with more net positive emotion words in the collected tweets received a higher (positive) daily sentiment score, while days with more net negative words received a negative daily sentiment score.
